### 2015-04-22 docs/parasite_mortatlity_manuscript.tex
Filled in results and added to discussion with overview of results

In this study, we propose a simple, novel method for estimating both whether or
not PIHM is occurring in a host-parasite system and the shape of the survival
function.  We compare the ability of our method to answer the two PIHM
questions given above to the results given by the Adjei Method.  We show that
our method is superior to the Adjei Method and that the Adjei Method has some
serious deficiencies when trying to answer either of the above questions.

@@

\subsection*{The Adjei method for estimating PIHM}

\subsection*{The likelihood method for estimating PIHM}

@@

We tested the likelihood method on the 6 datasets given in \cite{Crofton1971a}
and the 4 datasets given in \citep{Adjei1986} in which both papers reported
seeing PIHM in the respective populations.  We examined both whether or not the
Likelihood Method or the Adjei Method detected significant PIHM in the given
datasets and the resulting $LD_{50}$ predictions from both methods.  We then
compared these results to whether the authors stated that PIHM was a
significant in the given system.

@@

The precision of the $LD_{50}$ estimates for the Likelihood method decreased (increasing Coefficient of Variation) as sample size decreased for all parameter combinations we examined.  The $LD_{50}$ estimates from the Adjei Method showed a similar pattern, with large decreases in precision occurring for the steepest survival function across all values of $\mu_p$ (Figure X).

@@

@@

Parasite-induced host mortality is of substantial interest in many systems, but determining whether it is occurring given only observational data is notoriously difficult.  We show that the Adjei Method, the only currently proposed method to estimate the host survival function and the $LD_{50}$ from observational PIHM, has some serious methodological problems that result in biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.

To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
general method to determine both whether or not PIHM is occurring in a system
and the shape of the survival function.  We show that this method is
asymptotically unbiased when estimating the host-survival function for all of
the parameter space that we explored, but can produce seriously biased
estimates of the host survival function for sample sizes typically observed in
many host-parasite studies.  However, we found that the likelihood method
produces unbiased and precise estimates of the $LD_{50}$ for small, realistic
sample sizes.

Moreover, the likelihood method...

@@

While we have generalized and improved upon
the previously existing methods for answering questions about PIHM, we cannot belie the fact
that estimating PIHM from observational data alone is
ladened with assumptions and difficulties \citep{McCallum2000a}. The most fundamental
assumption of all methods for estimating PIHM is that the shape of the pre-
mortality host-parasite distribution is known. In the discrete case, this
distribution is negative binomial, while in the continuous case the distribution can be gamma or exponential (\citep{Ferguson2011}).  While there is substantial empirical and
theoretical evidence to justify the use of the negative binomial distribution
as the pre-mortality distribution for macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).  However, the critical assumption of the pre-mortality distribution is not that the processes leading the pre-mortality distribution generate a negative binomial distribution, but rather that the pre-mortality distribution is well-fit by a negative binomial. The extreme flexibility of the negative binomial distribution makes it a reasonable candidate distribution for the pre-mortality distributions.  Therefore, we do not see this assumption as central problem in any of the proposed methods.

@@

Related to the above result, all of these methods require that the post-mortality distribution be significantly different from a negative binomial distribution.  This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution (describe why?).  As many observed host-parasite distributions are not significantly different from a negative binomial distribution, there are limited cases where these methods can even be considered.

Finally, all of these methods assume that the truncation of a negative binomial
distribution is due to PIHM, but simulation studies by \citep{Anderson1982a}
show that a variety of other processes can lead to the truncation of a negative
binomial distribution such as within host parasite density-dependence, age-
dependent variation in host resistance and rate of infection
\citep{McCallum2000a}.  Therefore, even detecting ``significant'' PIHM in a
dataset does not mean that PIHM is cause of the truncation.

Given these numerous caveats, is there a place in parasitology for methods that
estimate PIHM from distributional data?  We are in agreement with
\cite{Lester1984} that, at the very least, methods for estimating PIHM can at
least provide some preliminary insight into whether or not PIHM is worth
looking into.  However, we stress that these methods should only be used as an
exploratory tool when assessing the role of PIHM in a system and potential
users should critically evaluate whether they think they have a large enough
sample sizes and appropriate host survival functions for the methods proposed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional substitute for field or laboratory experiments and an in
depth understanding of the natural history of the host-parasite system under
question.

@@

>>>>

In this study, we propose a novel method for estimating both whether or
not PIHM is occurring in a host-parasite system and the shape of the survival
function.  We compare the ability of our method to answer the two PIHM
questions given above to the results given by the Adjei Method.  We show that
our method is superior to the Adjei Method and that the Adjei Method has
serious deficiencies when trying to answer either of the above questions.

@@

\subsection*{The Adjei Method for estimating PIHM}

\subsection*{The Likelihood Method for estimating PIHM}

@@

We tested the ability of the Adjei Method and the Likelihood Method to identify
PIHM on 6 host-parasite datasets given in \cite{Crofton1971a} and 4 datasets
given in \cite{Adjei1986}. In the \cite{Crofton1971a} datasets, the host was
the snail \emph{Gammarus pulex} which acts as the intermediate host for the
acanthocephalan \emph{Polmorphus minutus}. In the \cite{Adjei1986} datasets,
the hosts were two species of lizard fish \emph{Saurida tumbil} and
\emph{Saurida undosquamis} that were infected by the cestode parasite
\emph{Callitetrarhynchus gracilis}.  Males and females of both fish species
were considered separately.

In both these studies, the authors reported PIHM and we test whether the Adjei
Method and or the Likelihood Method also predict PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  We then parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.

@@

The precision of the $LD_{50}$ estimates for the Likelihood Method decreased (increasing coefficient of variation) as sample size decreased for all parameter combinations we examined.  The $LD_{50}$ estimates from the Adjei Method showed a similar pattern, with large decreases in precision occurring for the steepest survival function across all values of $\mu_p$ (Figure X).

@@

Of the 10 datasets we considered, the previous authors claimed to detect PIHM
in 7 of them (Table \ref{table:pihm}).  The Likelihood Method parameterized
from the pre-mortality parameters of the Crofton Method detected significant
PIHM in 6 of these 7 datasets at a significance level of 0.05.  The only
dataset in which the the Likelihood Method parameterized from the Crofton
Method did not detect a significant effect of PIHM was the the Adjei dataset
for \emph{S. tumbil}.  For this case there was a marginally significant effect
of PIHM the likelihood ratio test giving $\chi^2_{df=2} = 5.34; p = 0.069$.

The Adjei Method detected significant PIHM in 9 of the 10 datasets given (Table \ref{table:pihm}).  This is consistent with the the previous results which show that the Adjei Method has a very high Type I error rate.  The Likelihood Method in which all parameters were jointly estimated only predicted significant PIHM in 4 of the 10 datasets.

Discuss LD50 values?

@@

Parasite-induced host mortality is of substantial interest in many systems, but
determining whether it is occurring given only observational data is
notoriously difficult.  We show that the Adjei Method, the only currently
proposed method to estimate the host survival function and the $LD_{50}$ from
observational PIHM, has some serious methodological problems that result in
biased estimates of the host survival function and the $LD_{50}$ even under the most idealistic conditions.  Moreover, we show that the Adjei Method has a seriously inflated Type I error rate, meaning it will often detect PIHM even when it is not present.

To attempt to ameliorate the flaws in the Adjei Method, we proposed a more
general method to determine both whether or not PIHM is occurring in a system
and the shape of the survival function.  We show that this method is
asymptotically unbiased when estimating the host-survival function for all of
the parameter space that we explored and we found that it
produces unbiased and precise estimates of the $LD_{50}$ for small, realistic
sample sizes.  Moreover, this novel method has reasonable type I and type II error rates when attempting to predict whether or not PIHM is occurring in a system.  However, we note that the Likelihood Method produces seriously biased
estimates of the host survival function ($a$ and $b$) for sample sizes typically observed in many host-parasite studies and should not be used to estimate the exact shape of the host survival function.

We fit both the Likelihood Method and the Adjei Method to empirical data to
determine whether they could detect PIHM that had been previously reported
based on visual assessments.  Consistent with our simulation results, we found
that the Adjei Method tended to detect PIHM where it had not been previously
reported, while the Likelihood Method's detection of PIHM was consistent with
previously PIHM in a given dataset. Taken together, these results suggest that the Adjei Method is a fundamentally
flawed method for detecting PIHM and describing features of the host survival
function and we recommend using the Likelihood Method for detecting PIHM and describing attributes of the host survival function.

@@

While we improved upon
the previously existing methods for answering questions about PIHM, we cannot belie the fact
that estimating PIHM from observational data alone is
ladened with assumptions and difficulties \citep{McCallum2000a}. The most fundamental
assumption of all methods for estimating PIHM is that the shape of the pre-
mortality host-parasite distribution is known. In the discrete case, this
distribution is negative binomial, while in the continuous case the distribution can be gamma or exponential (\citep{Ferguson2011}).  While there is substantial empirical and
theoretical evidence to justify the use of the negative binomial distribution
as the pre-mortality distribution for macroparasites across hosts \citep{Calabrese2011,Anderson1982a,Shaw1998}, it is widely recognized that different processes can lead to a variety of distributions of parasites across hosts (Wilber, Duerr etc).  However, the critical assumption of the pre-mortality distribution is not that the processes leading the pre-mortality distribution generate a negative binomial distribution, but rather that the pre-mortality distribution is well-fit by a negative binomial. The extreme flexibility of the negative binomial distribution makes it a reasonable candidate distribution for the pre-mortality distributions.  Therefore, we do not see this assumption as central problem in any of the proposed methods.

@@

Related to the above result, all of these methods require that the post-mortality distribution be significantly different from a negative binomial distribution.  This is necessary because none of the above methods will be able to detect PIHM if a negative binomial distribution is an adequate fit to the post-mortality distribution [describe why?].  As many observed host-parasite distributions are not significantly different from a negative binomial distribution, there are limited cases where these methods can even be considered.

Finally, all of these methods assume that the truncation of a negative binomial
distribution is due to PIHM, but previous studies have shownthat a variety of other processes can lead to the truncation of a negative
binomial distribution such as within host parasite density-dependence, age-
dependent variation in host resistance and rate of infection
\citep{McCallum2000a,Anderson1982a,Rousset1996}.  Therefore, even detecting ``significant'' PIHM in a
dataset does not mean that PIHM is cause of the truncation.

Given these numerous caveats, is there a place in parasitology for methods that
estimate PIHM from distributional data?  We are in agreement with
\cite{Lester1984} that, at the very least, methods for estimating PIHM can provide preliminary insight into whether or not PIHM is worth further exploration.  However, we stress that these methods should only be used as an
exploratory tool when assessing the role of PIHM in a system and potential
users should critically evaluate whether they think they have a large enough
sample sizes and appropriate host survival functions for the methods proposed
in this paper to be applicable.  Even if they are applicable, inferring PIHM
from distributional data is no substitute for field or laboratory experiments and an in
depth understanding of the natural history of the host-parasite system under consideration.

@@

\renewcommand{\arraystretch}{1.2}
\begin{sidewaystable}

    \caption{Comparison of the PIHM predictions of previously used host-parasite datasets to those given by the Adjei Method and the Likelihood Method. The first column specifies the identity of the dataset, the second column specifies whether or not the author indicated that PIHM was occurring in the system, the third column indicates whether or not the Likelihood Method with parameters from the Crofton Method detects significant PIHM, the third column indicates whether the Likelihood Method with jointly estimated pre-mortality parameters detects PIHM, and the final column indicates whether the Adjei Method with pre-mortality parameters estimated from the Crofton Method detects PIHM.  If a method detects significant PIHM the $LD_{50}$ is given is parentheses}

    \centering
    \begin{tabular}{l l p{3cm} p{3cm} l}

    \hline\hline
    Data Set & Author detected PIHM? & Likelihood Method w/ Crofton Parameters & Likelihood Method w/out Crofton Parameters & Adjei Method \\

    \hline
    Crofton, Station 1 & Yes & Yes (7.27) & Yes (8.00) & Yes (9.33) \\
    Crofton, Station 2 & Yes & Yes (6.92) & No & Yes (14.95)\\
    Crofton, Station 3 & Yes & Yes (5.93) & Yes (7.04) & Yes (5.98) \\
    Crofton, Station 4 & No & No & No & Yes (7.99) \\
    Crofton, Station 5 & No & No & No & Yes (10.58) \\
    Crofton, Station 6 & No & No & No & No \\
    Adjei, \emph{S. tumbil} female & Yes (5.7) & No  & No   & Yes (6.37) \\
    Adjei, \emph{S. tumbil} male & Yes (3.4) & Yes (3.42) & Yes (3.28) & Yes (3.66)  \\
    Adjei, \emph{S. undosquamis} female & Yes (3.2) & Yes (3.04) & No  & Yes (3.11) \\
    Adjei, \emph{S. undosquamis} male & Yes (1.8) & Yes (1.83) & Yes (2.34) & Yes (1.78) \\

    \end{tabular}
    \label{table:pihm}

\end{sidewaystable}
